---
title: "ENVS193_Final"
author: "Caleb Neely"
format: pdf
---

### Set Up
```{r}
library(tidyverse)
library(janitor)
library(gt)
library(DHARMa)

nests <- read_csv("nest_data_final.csv")
```


## Problem 1: Research writing

### a) Transparent statistical methods

Part 1: The co-worker likely used a logistic regression because the response variable is the probability of American Avocet microhabitat use (used vs. not used) and the predictor variable is distance to the water edge, which is continuous.

Part 2: The co-worker likely used a chi-square test of independence because both bird species and habitat type are categorical variables, and the goal is to determine whether the distribution of bird species differs among habitat types.


### b) Suggestions For Rewriting

Part 1: American Avocets were more likely to use some areas than others depending on their distance from the water's edge, suggesting that proximity to water influences microhabitat selection.

Part 2: The three bird species did not use habitats in the same way. Habitat use differed among species, indicating that each species may prefer different habitat types within the San Francisco Bay.


### c) Further Analysis

One additional variable that could influence the probability of American Avocet microhabitat use is water depth (cm), measured at each survey location using a meter stick or depth gauge. Water depth may affect habitat use because American Avocets forage by sweeping their bills through shallow water, so they may prefer areas with depths that maximize feeding efficiency.


## Problem 2 Data analysis

### a) Clean and Wrangle

```{r}
# Create a new cleaned dataset containing only the variables needed for the nest occurrence analysis
nests_clean <- nests |>
  # Keep only the response variable, tree height, and ant species columns
  select(case_control, height_cm, ant_species) |>
  # Keep only trees occupied by the three focal ant species
  filter(ant_species %in% c("AB", "BBR", "RRB")) |>
  # Recode and reorder ant species information
  mutate(
    # Convert the ant species codes into a factor and set the desired order
    ant_species = factor(
      ant_species,
      levels = c("BBR", "AB", "RRB"),
      labels = c(
        "C. sjostedti",
        "C. mimosae",
        "C. nigriceps"
      )
    ),
     # Create a new column containing the full scientific name of each ant species
    ant_species_full = case_when(
      ant_species == "C. sjostedti" ~ "Crematogaster sjostedti",
      ant_species == "C. mimosae" ~ "Crematogaster mimosae",
      ant_species == "C. nigriceps" ~ "Crematogaster nigriceps"
    )
  )
# Display the structure of the cleaned dataset to verify variable types and dimensions
str(nests_clean)
# Display 10 randomly selected rows from the cleaned dataset
nests_clean |>
  slice_sample(n = 10)
```


### b) Response variable

The response variable is nest occurrence, where a value of 1 indicates that a bird nest was present in a tree and a value of 0 indicates that no bird nest was present. Biologically, this variable represents whether a tree was selected and used as a nesting site by birds.


### c) Purpose of Study

Tree height:

Tree height is a continuous variable measured in centimeters. As tree height increases, the probability of nest occurrence may increase because taller trees can provide better nesting locations and may be more visible or accessible to birds while also reducing the risk of some predators. This information is discussed in the Methods section, “Nest Site Selection” subsection, where tree characteristics were measured, and is supported by the Results section, where tree height was included as a predictor of nest occurrence.

Acacia ant species:

Acacia ant species is a categorical variable with three levels: C. sjostedti, C. mimosae, and C. nigriceps. The probability of nest occurrence may differ among ant species because birds preferentially nest in trees occupied by more aggressive ant species that defend trees against browsing mammals and potential nest predators. This information is described in the Introduction (final paragraph) and supported in the Discussion section, where the authors explain that birds selected trees inhabited by the more aggressive ant species.


### d) Table of Models

```{r}
# Create a table that summarizes the two candidate logistic regression models
model_table <- tibble(
  # Give each model a descriptive name
  `Model Name` = c("Additive Model", "Interactive Model"),
  # Identify the response variable for both models
  `Response Variable` = c(
    "Nest occurrence",
    "Nest occurrence"
  ),
  # Describe the predictors included in each model
  `Predictors` = c(
    "Tree height + acacia ant species",
    "Tree height + acacia ant species + tree height x acacia ant species"
  ),
  # State whether each model includes an interaction term
  `Interaction Term` = c(
    "No",
    "Yes"
  )
)
# Display the model table
model_table |>
  gt() |>
  tab_header(
    title = "Candidate Logistic Regression Models"
  )
```


### e) Run the models

```{r}
# Run the additive logistic regression model
additive_model <- glm(
  case_control ~ height_cm + ant_species,
  data = nests_clean,
  family = binomial
)
# Run the interaction logistic regression model
interaction_model <- glm(
  case_control ~ height_cm * ant_species,
  data = nests_clean,
  family = binomial
)
```


### f) Select the best model

```{r}
AIC(additive_model, interaction_model)
```

The best model as determined by Akaike’s Information Criterion (AIC) was the logistic regression model predicting nest occurrence from tree height, acacia ant species, and the interaction between tree height and acacia ant species. This model had a lower AIC value (237.804) than the model containing only the additive effects of tree height and acacia ant species (258.743), indicating stronger support for the interaction model.


### g) Check the diagnostics

```{r}
# Simulate residuals from the selected model
interaction_residuals <- simulateResiduals(
  fittedModel = interaction_model
)

# Diagnostic plots
plot(interaction_residuals)

# Test for uniformity
testUniformity(interaction_residuals)

# Test for dispersion
testDispersion(interaction_residuals)

# Test for outliers
testOutliers(interaction_residuals)
```






